\jaxcosmo: An End-to-End Differentiable and GPU Accelerated Cosmology Library

### 2022-11-14 paper_OJA.tex
Review SVI alone result according to new MVN convergence.

Here is the simple way \numpyro\ deals with SVI initialisation and optimisation\footnote{See also listing in Appendix \ref{app-DESY1}) where the \texttt{model} is described and the \texttt{cl\_obs} are the observed $C_\ell$.}
\begin{lstlisting}[language=iPython]
import numpyro.infer.autoguide as autoguide
from numpyro.infer import SVI, Trace_ELBO
from numpyro.optim import Adam
# choose a SVI guide 
guide = autoguide.AutoMultivariateNormal(model,
        init_loc_fn=numpyro.infer.init_to_median())
# SVI optimisation 
optimizer = numpyro.optim.Adam(1e-3)
svi = SVI(model_spl, guide,optimizer,loss=Trace_ELBO(num_particles=10))
svi_result = svi.run(jax.random.PRNGKey(0),20_000, cl_obs)
#To get samples from SVI approximated posterior 
samples = guide.sample_posterior(jax.random.PRNGKey(1), svi_result.params, sample_shape=(100_000,))
\end{lstlisting}
%
\JEC{Once the optimisation is done on gets $z$ i.i.d. samples from the $q(z,\lambda^\ast)$ distribution  applying the inverse of the inverse of the $S_{\lambda^\ast}$ and $T$ transformations. Using the same DES-Y1 simulation as in previous section, we use both a MultiVariate Normal distribution (MVN) and a Block Nueral Autoregressive Flow (BNAF) \citep{pmlr-v115-de-cao20a} as \textit{guides} to approximate the true posterior (\textit{model}). The SVI optimization has been performed with the Adam optimizer \citep{2014arXiv1412.6980K} and a learning rate set to $10^{-3}$. We have stopped the optimization after 20,000 steps to ensure a good convergence of the ELBO loss, but no particular tuning of the optimizer has been undertaken to decrease the number of steps. It takes about 2 hours on the NVidia V100 GPU}.  \FrL{Arf... 2 hours !}

\JEC{
In the figure \ref{fig_cobaya_SVIs}, we compare the contours obtained with \texttt{Cobaya} as in figure \ref{fig_cobaya_NUTS_SVI}) and those obtained with the MVN and BNAF guides for the SVI.  As mentioned in \cite{NEURIPS2020_7cac11e2} one challenge with variational inference is assessing how close the variational approximation $q(z,\lambda^\ast)$ is to the true posterior distribution. It is not in the scope of this article to elaborate a statistical diagnostics, rather we show that with SVI, one can get a quick estimate of the posterior which can serve as front-end to a NUTS sampler. This techniques is described in the next section.}

%
\begin{figure*}
\centering
\includegraphics[width=1.5\columnwidth]{figures/fig_Cobaya_SVIs_v1.pdf}
\caption{Same configuration as in Figure \ref{fig_cobaya_NUTS_SVI} but using sampling of approximated perameter posteriorf using SVI with MVN and BNAF guides compared to \texttt{Cobaya} sampling.}
\label{fig_cobaya_SVIs}
\end{figure*}
%
%
\subsubsection{Neural Transport}
\label{sec-Neural-Reparametrisation}
%
If the SVI method can be used as is to get $z$ i.i.d. samples from the $q(z,\lambda^\ast)$ distribution as shown on the previous section, the \textit{Neural Transport MCMC} method \citep{2014arXiv1412.5492P,2019arXiv190303704H} is an efficient way to boost HMC efficiency, especially in target distribution unfavourable geometry where for instance the leapfrog integration algorithm has to face squeezed join distributions for a subset of variables. From SVI, one obtains a first approximation of the target distribution and we show that the change-of-variable $z=F_\lambda(\zeta)$ (eg. $F_\lambda=T^{-1}\circ S^{-1}_\lambda$) is such that
\begin{equation}
q(z;\lambda) \rightarrow q(\zeta;\lambda) \bydef q(F_\lambda(\zeta)) |J_{F_\lambda}(\zeta)|
\end{equation}
which is evaluated with optimal $\lambda^\ast$ according to maximization of the ELBO. So, one can use a HMC sampler (eg. NUTS) based on $p(\zeta;\mathcal{D})$ distribution, initialized with $\zeta$ samples from $q(\zeta;\lambda^\ast)$, to get a Markov Chain of $N$ samples $(\zeta_i)_{i<N}$. Then, from the transformation  $z_i=F_{\lambda^\ast}(\zeta_i)$ one finally obtain a Marcov Chain with $(z_i)_{i<N}$ samples. 
\begin{figure*}
\centering
\includegraphics[width=1.5\columnwidth]{figures/fig_Cobaya-NUTS-SVI200_bis_v1.pdf}
\caption{Same configuration as in Figure \ref{fig_cobaya_NUTS_SVI} but for the highly correlated five lens galaxy bias.}
\label{fig_cobaya_NUTS_SVI_bis}
\end{figure*}
The \textit{Neural Transport} is accessible throw the  \verb|NeuTraReparam| in \numpyro\ library as followed:
\begin{lstlisting}[language=iPython]
from numpyro.infer.reparam import NeuTraReparam
# Reparametrisation of model to feed NUTS sampler
neutra = NeuTraReparam(guide, svi_result.params)
neutra_model = neutra.reparam(model_spl) 
\end{lstlisting}
%
We have used the MVN guided SVI described in the previous section even if it may be not the best choice. The NUTS sampler has run 1 chain of 200 samples (``low'' schema), 1 chain with 1000 samples and 10 chains  with 1000 samples each (each setup starts all chains with 200 samples for initialisation). All NUTS sampling as been performed with dense mass matrix optimisation without special block structuring and with \texttt{max\_tree\_depth=5} which is different than the default NUTS setting described in section \ref{sec-NUTS}. The elapse time for each of the 3 setups is 50 minutes, 150 minutes and 5 hours, respectively. Naturally, more samples lead to better contour precision. But what is illustrative is the ``low'' schema results as shown in the Figures \ref{fig_cobaya_NUTS_SVI} and \ref{fig_cobaya_NUTS_SVI_bis}  to be compared to \texttt{Cobaya} and NUTS results presented in section \ref{sec-NUTS}. Notice the good results even for the highly correlated lens bias parameters. It is noticeable that running NUTS with the same settings  but without the SVI Neural Transport phase has demonstrated a rather poor behaviour with only 200 samples.

>>>>

Here is the simple way \numpyro\ deals with SVI initialisation and optimisation\footnote{See also listing in Appendix \ref{app-DESY1}) where the \texttt{model} is described and the \texttt{cl\_obs} are the observed $C_\ell$.}
\begin{lstlisting}[language=iPython]
import numpyro.infer.autoguide as autoguide
from numpyro.infer import SVI, Trace_ELBO
from numpyro.optim import Adam
# choose a SVI guide 
guide = autoguide.AutoMultivariateNormal(model,
        init_loc_fn=numpyro.infer.init_to_median())
# SVI optimisation 
optimizer = numpyro.optim.Adam(1e-3)
svi = SVI(model_spl, guide,optimizer,loss=Trace_ELBO(num_particles=10))
svi_result = svi.run(jax.random.PRNGKey(0),20_000, cl_obs)
#To get samples from SVI approximated posterior 
samples = guide.sample_posterior(jax.random.PRNGKey(1), svi_result.params, sample_shape=(100_000,))
\end{lstlisting}
%
\JEC{Once the optimisation is done on gets $z$ i.i.d. samples from the $q(z,\lambda^\ast)$ distribution  applying the inverse of the inverse of the $S_{\lambda^\ast}$ and $T$ transformations. Using the same DES-Y1 simulation as in previous section, we use both a MultiVariate Normal distribution (MVN) and a Block Nueral Autoregressive Flow (BNAF) \citep{pmlr-v115-de-cao20a} as \textit{guides} to approximate the true posterior (\textit{model}). The SVI optimization has been performed with the Adam optimizer \citep{2014arXiv1412.6980K} and a learning rate set to $10^{-3}$. We have stopped the optimization after 20,000 (30,000) steps to ensure a stable ELBO loss convergence when using the BNAF (MVN) guides, and no particular tuning of the optimizer has been undertaken as for instance a learning rate scheduling. It takes about 2 or 3 hours on the NVidia V100 GPU scaling according to the number of steps}.  \FrL{Arf... 2 hours !}

\JEC{
In the figure \ref{fig_cobaya_SVIs}, we compare the contours obtained with \texttt{Cobaya} as in figure \ref{fig_cobaya_NUTS_SVI}) and those obtained with the MVN and BNAF guided SVI.  As mentioned in \cite{NEURIPS2020_7cac11e2} one challenge with variational inference is assessing how close the variational approximation $q(z,\lambda^\ast)$ is to the true posterior distribution. It is not in the scope of this article to elaborate a statistical diagnostics, rather we show that both guided SVI exhibit rather similar contours and both estimate are close to the Cobaya posterior sampling in a short time scale. This SVI posterior estimates can serve as front-end to a NUTS sampler as described in the next section.} 
%
\begin{figure*}
\centering
\includegraphics[width=1.5\columnwidth]{figures/fig_Cobaya_SVIs_v1.pdf}
\caption{Same configuration as in Figure \ref{fig_cobaya_NUTS_SVI} but using sampling of approximated posterior using SVI with MVN and BNAF guides (see text) compared to \texttt{Cobaya} sampling.}
\label{fig_cobaya_SVIs}
\end{figure*}
%
\subsubsection{Neural Transport}
\label{sec-Neural-Reparametrisation}
%
If the SVI method can be used as is to get $z$ i.i.d. samples from the $q(z,\lambda^\ast)$ distribution as shown on the previous section, the \textit{Neural Transport MCMC} method \citep{2014arXiv1412.5492P,2019arXiv190303704H} is an efficient way to boost HMC efficiency, especially in target distribution unfavourable geometry where for instance the leapfrog integration algorithm has to face squeezed join distributions for a subset of variables. From SVI, one obtains a first approximation of the target distribution and we show that the change-of-variable $z=F_\lambda(\zeta)$ (eg. $F_\lambda=T^{-1}\circ S^{-1}_\lambda$) is such that
\begin{equation}
q(z;\lambda) \rightarrow q(\zeta;\lambda) \bydef q(F_\lambda(\zeta)) |J_{F_\lambda}(\zeta)|
\end{equation}
which is evaluated with optimal $\lambda^\ast$ according to maximization of the ELBO. So, one can use a HMC sampler (eg. NUTS) based on $p(\zeta;\mathcal{D})$ distribution, initialized with $\zeta$ samples from $q(\zeta;\lambda^\ast)$, to get a Markov Chain of $N$ samples $(\zeta_i)_{i<N}$. Then, from the transformation  $z_i=F_{\lambda^\ast}(\zeta_i)$ one finally obtain a Marcov Chain with $(z_i)_{i<N}$ samples. 
\begin{figure*}
\centering
\includegraphics[width=1.5\columnwidth]{figures/fig_Cobaya-NUTS-SVI200_bis_v1.pdf}
\caption{Same configuration as in Figure \ref{fig_cobaya_NUTS_SVI} but for the highly correlated five lens galaxy bias.}
\label{fig_cobaya_NUTS_SVI_bis}
\end{figure*}
The \textit{Neural Transport} is accessible throw the  \verb|NeuTraReparam| in \numpyro\ library as followed:
\begin{lstlisting}[language=iPython]
from numpyro.infer.reparam import NeuTraReparam
# Reparametrisation of model to feed NUTS sampler
neutra = NeuTraReparam(guide, svi_result.params)
neutra_model = neutra.reparam(model_spl) 
\end{lstlisting}
%
We have used the MVN guided SVI described in the previous section. The NUTS sampler has run 1 chain of 200 samples (``low'' schema), 1 chain with 1000 samples and 10 chains  with 1000 samples each (each setup starts all chains with 200 samples for initialisation). All NUTS sampling as been performed with dense mass matrix optimisation without special block structuring and with \texttt{max\_tree\_depth=5} which is different than the default NUTS setting described in section \ref{sec-NUTS}. The elapse time for each of the 3 setups is 50 minutes, 150 minutes and 5 hours, respectively. Naturally, more samples lead to better contour precision. But what is illustrative is the ``low'' schema results as shown in the Figures \ref{fig_cobaya_NUTS_SVI} and \ref{fig_cobaya_NUTS_SVI_bis}  to be compared to \texttt{Cobaya} and NUTS results presented in section \ref{sec-NUTS}. Notice the good results even for the highly correlated lens bias parameters. It is noticeable that running NUTS with the same settings  but without the SVI Neural Transport phase has demonstrated a rather poor behaviour with only 200 samples.
